---
title: "practice"
output: html_document
date: "2023-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Class 01: Quick start

### Slides' Questions 1

- The airquality is a dataset available in R
```{r airquality}
summary(airquality)
```
- How many variables there are in the data?
Using help, we can find out how many variable the data has
```{r}
help("airquality")
```
It has 6 variables, Ozone, Solar, Wind, Temp, Month, Day

- Define an R object which contain the information about the wind speed.

```{r}
wind_speed <- airquality$Wind
```
- Calculate the mean, and variance for the wind speed.
```{r}
print(mean(wind_speed))
print(var(wind_speed))
```
### Slides' Questions 2
- The ToothGrowth is a dataset available in R. Use help for more details
```{r}
help("ToothGrowth")
```
This dataset is data about the effect of Vitamin C on Tooth Growth in Guinea Pigs. Data has three variables, len, supp, and dose

- The response variable is the Tooth length. Test if the Supplement type has an effect on the tooth length
```{r}
help(t.test)
t.obj <- t.test(len ~ supp, data = ToothGrowth, var.equal = T)
print(t.obj)
```
## Slides' Question 3

- Use the ToothGrowth data, and produce an histogram


```{r pressure, echo=FALSE}
help(hist)
hist(ToothGrowth$len, xlab="x", main="Tooth Length", col="green")
```
## Class 01: Introduction to R

- Create a dataframe
```{r}
letters = c("A", "B", "C", "D", "E", "F")
numbers = c(100, 99, 105, 35, 0, 250)
test <- data.frame(letters, numbers)
```
- Create a file with data and read it
```{r}
fake_data <- read.table("ex1_data.txt")
```

- print your name x times
```{r}
x <- 10
for (i in 1:10 ){
  print("my name")
}
```

- Write a function which receive a numerical vector as an input and calculate the mean of the vector.

```{r}
my_mean <- function(vector) {
  return (mean(vector))
}
print(mean(c(1, 2, 3)))
```