---
title: "PracticalSession1"
output: html_document
date: "2023-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Questions
## Question 1
- Draw a sample of size 10 from N(0, 1), and get mean, var.
```{r}
x <- rnorm(10, 0, 1)
Stat <- c("mean", "var", "median")
Value <- c(mean(x), var(x), median(x))
object <- data.frame(Stat, Value)
print(object)
```
## Question 2
- Draw a random sample of size 100 from t(3)
- Produce a histogram
```{r}
x <- rt(100, 3)
hist(x, main="histogram of x", xlab="x", xlim=c(-5, 10), col="red", nclass=8)
```
## Question 3
1. Define a vector x =(1,..10)
2. Define an object y = x^2
3. Plot x vs y

```{r}
x <- 1:10
y <- x ^2
plot(x, y)
```
## Question 4
- We need cars data
- print speed, stopping distance
- plot speed vs distance
- spearman corr speed and distance
- min, max, median
- y = (median(distance) + 3) * 10, get the mean of y
```{r}
speed = cars$speed
distance = cars$dist
print("speed")
print(speed)
print("distance")
print(distance)
plot(speed, distance)
print("correlation")
print(cor(speed, distance, method="pearson"))
print("min")
print(min(speed))
print("max")
print(max(speed))
print("median")
print(median(speed))
y = (median(distance) + 3) * 10
print("median of y")
print(median(y))
```
## Question 5
Produce a specific dataframe
```{r}
x <- c(1, 3, 5, 7)
y <- c("one", "three", "five", "seven")
df <- data.frame(x, y)
print(df)
```

## Question 6
From airquality dataset
- How many observations and how many variables
```{r}
help(airquality)
print("153 observations and 6 variables")
```
- Define an object x that contains temperature
- Define an object Month
- Boxplot
```{r}
x = airquality$Temp
Month = airquality$Month
print(x)
print(Month)
print(split(x, Month))
boxplot(split(x, Month))
```

## Question 7
1. y with ozone leve
2. mean, median, min, max, std in a table
3. graphs
```{r}
y = airquality$Ozone
x = c("Mean", "Median", "Min", "Max", "Std")
w = c(mean(y, na.rm=T), median(y, na.rm=T), min(y, na.rm=T), max(y, na.rm=T), sd(y, na.rm=T))
df = data.frame(x, w)
print(df)
par(mfrow = c(1, 2))
hist(y, main = "Histogram for Ozone Level")
boxplot(y, main = "Boxplot for Ozone Level")
```

## Question 8
- Define R as wind/temp
- plot
```{r}
R = airquality$Wind / airquality$Temp
plot(airquality$Day, R)
```

## Question 9
- From last question add a red line with the median
- Change labels

```{r}
R = airquality$Wind / airquality$Temp

plot(airquality$Day, R, xlab="Day in the Month", ylab = "Ratio(Wind/Temp)")
abline(h = median(R), col = "red")
```

## Question 10

Using for loop
- draw 1000 samples of size 31 from B(1, 0.5)
- Calculate the mean for each sample 
- Plot a histogram of the sample mean

```{r}
N = 1000
m.x = c(1:N)
for (i in 1:N) {
  sample <- rbinom(31, 1, 0.5)
  m.x[i] <- mean(sample)
}
hist(m.x)
```



