---
title: "Practical Session2a"
output: pdf_document
author: Juan Vanegas
date: "2023-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Information
- The practical session consists of 5 questions in which you are asked to conduct an analysis of a dataset or to use R for simple calculations. Datasets are available in R as data frames.
- Your output should consists a PDF file which contains the results and R code.
- Solutions will be available online in BB in a later stage (you will receive an email about this via BB).

## Question 1

In this question we focus on the airquality data which available as data frame in R. Use help(airquality) to
get information about the data.
1. Define an R object which contain the information about the wind speed.
2. Calculate the mean, median and variance for the wind speed.
3. Produce the figure below.
```{r}
windspeed <- airquality$Wind
sprintf("mean: %f", mean(windspeed))
sprintf("var: %f", var(windspeed))
sprintf("median: %f", median(windspeed))
```
```{r}
par(mfrow=c(1,2))
hist(windspeed, xlab = "X", main = "Histogram of wind speed")
boxplot(windspeed, col=2)
```
## Question 2
In this question we focus on the ToothGrowth data which available as data frame in R. Use
help(ToothGrowth) to get information about the data.
1. The response variable is the Tooth length.
2. Test if the Supplement type has an effect on the tooth length.
```{r}
head(ToothGrowth)
```
```{r}
dim(ToothGrowth)
```
```{r}
t.test(len~supp, var.equal=TRUE, data=ToothGrowth)
```
```{r}
boxplot(ToothGrowth$len~ToothGrowth$supp, xlab="Group", ylab = "length")
```
## Question 3

For the ToothGrowth data, produce an histogram for the tooth length (by group) with the following structure.

```{r}
par(mfrow=c(2,1))
hist(ToothGrowth$len[ToothGrowth$supp == "VC"], col=2, xlim=c(0, 40))
hist(ToothGrowth$len[ToothGrowth$supp == "OJ"], col=3, xlim=c(0, 40))
```
## Question 4
Create the following data frame
```{r}
Name <- c("A", "B")
Gender <- c("M", "M")
Height <- c(180, 170)
data.frame(Name, Gender, Height)
```

## Question 5
In this question we focus on the chickwts data which available as data frame in R. Use help(chickwts) to get
information about the data.
1. Create a new data frame which contains data only of the horsebean and the soybean groups. How many observations are included in the new dataset.
2. calculate the mean and standard deviation of the weight by group.

```{r}
newdata <- chickwts[chickwts$feed %in% c("horsebean", "soybean"), ]
dim(newdata)
```
```{r}
tapply(newdata$weight, newdata$feed, mean)
tapply(newdata$weight, newdata$feed, sd)

```


